The {TestIdeals} package for \emph{Macaulay2}

### 2019-04-29 Fsing/TestIdealsPaper/TestIdeals.tex
More work on examples in paper

\medskip
{\small
\setstretch{.67}
\begin{MyVerbatim}
i1 : R = ZZ/5[x,y,z];

@@

We can also describe Frobenius roots as follows:  In a (sufficiently local) regular ring, we have an identification of $R$ with its canonical module $\omega_R$.
On the other hand, the Grothendieck dual to the $e$-iterated Frobenius map\footnote{In a reduced ring $R$, it is often convenient to express the Frobenius endomorphism $R \to R$ as the inclusion $R \hookrightarrow R^{1/p}$, so that the source and target  are distinguished.} $R \to R^{1/p^e}$ provides a map
\begin{equation}
\label{eq.DualToFrobenius}
T : \omega_{R^{1/p^e}} \to \omega_R,
\end{equation}
which using the identification $R \cong \omega_R$ gives us a map
\[
T : R^{1/p^e} \to R.
\]
It is not difficult to prove that
$T(I^{1/p^e})=I^{[1/p^e]}$,
where $I^{1/p^e} \subseteq R^{1/p^e}$ is simply the ideal of $p^e$-th roots of the elements of $I$.
In the case that
\[
R = \bF_{p^e}[x_1, \ldots, x_d],
\]
$R^{1/p^e}$ is a free $R$-module with basis consisting of the monomials
\[
\mathbf{x}^{\boldsymbol{\lambda}/p^e}= x_1^{\lambda_1/p^e} \cdots x_d^{\lambda_d/p^e},
\]
where $0 \leq \lambda_i \leq p^e-1$.
Furthermore, the map $T : R^{1/p^e} \to R$ is simply the projection defined as follows:
\[
T(x_1^{\lambda_1/p^e} \cdots x_d^{\lambda_d/p^e}) = \left\{ \begin{array}{rl} 1 &\text{if } \lambda_i = p^{e}-1 \text{ for all }i \\ 0 & \text{otherwise}  \end{array} \right.
\]
Using this, it is not difficult to see that if
\begin{equation}
\label{eqn1}
f^{1/p^e} = \sum_{\boldsymbol{\lambda}} f_{\boldsymbol{\lambda}} \mathbf{x}^{\boldsymbol{\lambda} / p^e},
\end{equation}
where $\boldsymbol{\lambda}$ runs over the tuples $(\lambda_1, \dots, \lambda_d)$ with $0 \leq \lambda_j \leq p^e-1$, and $f_{\boldsymbol{\lambda}} \in R$, then
$(f)^{[1/p^e]} = T( (f)^{1/p^e})$ is the ideal generated by the $f_{\boldsymbol{\lambda}}$.
We can then compute $I^{[1/p^e]}$ for more general $I$ by linearity: if $I=(f_1,\ldots,f_m)$, then $I^{[1/p^e]}= \sum(f_i)^{[1/p^e]}$.

@@

The computation of Frobenius roots of ideals reduces to the case of principal ideals, and its complexity grows linearly with the number of generators of the ideal.
Furthermore the calculation of  $(f)^{[1/p^e]}$ reduces to finding the terms in the right hand side of (\ref{eqn1}), which essentially amounts to
taking each term in $f$, computing the $p^e$-th root of the coefficient in $\mathbb{K}$, and dividing the monomial exponent vector by $p^e$ with remainder.
Hence the complexity of computing $(f)^{[1/p^e]}$ is proportional to the number of terms in $f$, and is independent of its degree.
We emphasize that the calculation of Frobenius roots \emph{does not involve the calculation of Gr\"obner bases.}

@@

\medskip
{\small
   \setstretch{.67}
\begin{MyVerbatim}
i1 : S = ZZ/5[x,y,z];

@@

\medskip
{\small
   \setstretch{.67}
\begin{MyVerbatim}
i1 : p = 3;

@@

\medskip
{\small
   \setstretch{.67}
\begin{MyVerbatim}
i8 : S = ZZ/3[a..f,x,y];

@@

\medskip
{\small
   \setstretch{.67}
\begin{MyVerbatim}
i1 : R = ZZ/2[a,b,c,d];

@@

\medskip
{\small\setstretch{.67}
\begin{MyVerbatim}
i7 : U = matrix entries E2;

@@

i8 : A = matrix entries relations source E2;

             2       3
o8 : Matrix R  <--- R

i9 : frobeniusRoot(1, U)

o9 = {-2} | 1 0 0 |
     {0}  | 0 d a |

              2       3
o9 : Matrix R  <--- R
\end{MyVerbatim}
}\medskip

@@

\medskip
{\small\setstretch{.67}
\begin{MyVerbatim}
i10 : V = ascendModule(1, A, U)

o10 = | 0   a  bc  |
      | c+d d2 a2d |

              2       3
o10 : Matrix R  <--- R
\end{MyVerbatim}
}\medskip

@@

\medskip
{\small\setstretch{.67}
\begin{MyVerbatim}
i1 : R = ZZ/7[x,y,z]/(x^3 + y^3 + z^3);

@@

\medskip
{\small\setstretch{.67}
\begin{MyVerbatim}
i1 : R = ZZ/7[x,y,z]/((x - 1)^5 + (y + 1)^5 + z^5);

@@

\medskip
{\small\setstretch{.67}
\begin{MyVerbatim}
i1 : R = ZZ/5[x,y,z]/(x^2 + y*z);

@@

\medskip
{\small\setstretch{.67}
\begin{MyVerbatim}
i1 : R = ZZ/5[x,y];

@@

\medskip
{\small\setstretch{.67}
\begin{MyVerbatim}
i1 : R = ZZ/7[x,y,z]/((x - 1)^3 + (y + 1)^3 + z^3);

@@

\medskip
{\small\setstretch{.67}
\begin{MyVerbatim}
i1 : S = ZZ/3[a,b,c,d,t];

@@

\medskip
{\small\setstretch{.67}
\begin{MyVerbatim}
i1 : R = ZZ/5[x,y,z]/(x^4 + y^4 + z^4);

@@

\medskip
{\small\setstretch{.67}
\begin{MyVerbatim}
i1 : R = ZZ/5[x,y,z]/(y*z, x*z, x*y);

@@

\medskip
{\small\setstretch{.67}
\begin{MyVerbatim}
i1 : R = ZZ/2[a..e];

@@

\medskip
{\small\setstretch{.67}
\begin{MyVerbatim}
i1 : T = ZZ/7[x,y];

@@

\medskip
{\small\setstretch{.67}
\begin{MyVerbatim}
i7 : toString QGorensteinGenerator(1, R)

@@

\medskip
{\small
\setstretch{.67}
\begin{MyVerbatim}
i2 : T = ZZ/3[x,y];

i3 : S = ZZ/3[a,b,c,d];

i4 : f = map(T, S, {x^3, x^2*y, x*y^2, y^3});

o4 : RingMap T <--- S

i5 : I = ker f;

i6 : R = S/I;

i7 : testIdeal R

o7 = ideal 1

o7 : Ideal of R
\end{MyVerbatim}
}\medskip

@@

Again, consider the maps $T^e : \omega_{R^{1/p^e}} \to \omega_R$ we have considered throughout this section.  It is a theorem of Hartshorne-Speiser, Lyubeznik, and Gabber \cite{HartshorneSpeiserLocalCohomologyInCharacteristicP,LyubeznikFModulesApplicationsToLocalCohomology,Gabber.tStruc} that the descending images
\begin{equation}\label{eq: descending images}
\omega_R \supseteq T(\omega_{R^{1/p}}) \supseteq \dots \supseteq T^e(\omega_{R^{1/p^e}}) \supseteq T^{e+1}(\omega_{R^{1/p^{e+1}}}) \supseteq \cdots
\end{equation}
stabilize for $e \gg 0$.
The function \texttt{FPureModule} computes the stable submodule in this chain, called the \emph{HSLG module}, and returns a list containing the following items:
\begin{enumerate}[(1)]
   \item The HSLGmodule;
   \item The canonical module, embedded (non-uniquely) as an ideal;
   \item The element $u$ representing the map on the canonical module (see
   \autoref{subsec.DualToFrobeniusOnQuotientRings});
   \item The value of $e > 0$ at which the descending images in \eqref{eq: descending images} stabilize.  This is sometimes also called the HSLG number of the canonical module as a Cartier module.
\end{enumerate}

\medskip
{\small
\setstretch{.67}
\begin{MyVerbatim}
i2 : R = ZZ/3[x,y,z]/ideal( x^3 + y^4 + z^5 );

i3 : L = FPureModule R;

i4 : L#0

                       2        2   3
o4 = ideal (y*z, x*z, y , x*y, x , z )

@@

i5 : L#1

o5 = ideal 1

o5 : Ideal of R

i6 : L#3

o6 = 1i2 : R = ZZ/3[x,y,z]/ideal( x^3 + y^4 + z^5 );

i3 : L = FPureModule R;

i4 : L#0

                       2        2   3
o4 = ideal (y*z, x*z, y , x*y, x , z )

i5 : L#1

o5 = ideal 1

o5 : Ideal of R

i6 : L#3

o6 = 1
\end{MyVerbatim}
}\medskip

@@

\medskip
{\small
\setstretch{.67}
\begin{MyVerbatim}
i2 : R = ZZ/2[x,y,z];

i3 : f = x^3 + y^2*z + y*z^2;

i4 : frobeniusPower(1/2, ideal f)

i5 : u=f^3;

i6 : frobeniusPower(1/4, ideal u)

o6 = ideal (z, y, x)

@@

o7 = 2
\end{MyVerbatim}
}\medskip

@@

\medskip
{\small
\setstretch{.67}
\begin{MyVerbatim}
i2 : R = ZZ/11[x,y,z];

i3 : f = y^2*z^3 - x^5 - 2*z^5;

i4 : frobeniusPower(1/11, ideal f^10)

             2        3
o4 = ideal (z , x*z, x )

@@

             2        3
o5 = ideal (z , x*z, x )

o5 : Ideal of R

i6 : ((FPureModule(1, f))#3) + 1

o6 = 2
\end{MyVerbatim}
}\medskip

@@

\medskip
{\small
\setstretch{.67}
\begin{MyVerbatim}
i2 : R = ZZ/3[u,v];

i3 : u = u^2*v^2;

i4 : compatibleIdeals u

o4 = {ideal v, ideal (v, u), ideal u}

o4 : List
\end{MyVerbatim}
}\medskip

@@

\medskip
{\small
\setstretch{.67}
\begin{MyVerbatim}
i2 : R = ZZ/2[a..f];

i3 : E = matrix {{a, b, b, e}, {d, d, c, a}};

             2       4
o3 : Matrix R  <--- R

i4 : I = minors(2, E);

i5 : S = R/I;

i6 : isCohenMacaulay S

@@

\medskip
{\small
\setstretch{.67}
\begin{MyVerbatim}
i7 : omega = canonicalIdeal S

o7 = ideal (e, d, a)

o7 : Ideal of S

i8 : omega = substitute(omega, R) + I;

@@

i9 : u = intersect((frobenius I):I, (frobenius omega) : omega);

o9 : Ideal of R

i10 : u = compress((gens u) % gens(frobenius I))

o10 = | a3bc+a3bd+a2cde+abcde+abd2e+b2d2e+cd2e2+d3e2 |

              1       1
o10 : Matrix R  <--- R

i11 :  u = first first entries u

       3       3       2                         2     2 2       2 2    3 2
o11 = a b*c + a b*d + a c*d*e + a*b*c*d*e + a*b*d e + b d e + c*d e  + d e

o11 : R
\end{MyVerbatim}
}
\medskip

@@

\medskip
{\small
\setstretch{.67}
\begin{MyVerbatim}
L = compatibleIdeals u;

i13 : print \ L;
       3       3       2                         2     2 2       2 2    3 2
ideal(a b*c + a b*d + a c*d*e + a*b*c*d*e + a*b*d e + b d e + c*d e  + d e )
               2
ideal (a + b, a  + d*e)
ideal (e, a, d)
ideal (e, d, a, c)
ideal (e, d, c, b, a)
ideal (e, d, b, a)
ideal (e, b, a)
ideal (e, b, a, c + d)
                      2
ideal (c + d, a + b, b  + d*e)
ideal (d, a)
ideal (d, a, c)
ideal (d, c, b, a)
ideal (d, b, a)
\end{MyVerbatim}
}
\medskip

@@

\medskip
{\small
\setstretch{.67}
\begin{MyVerbatim}
 print \ unique apply(L, J -> J : omega);
       3       3       2                         2     2 2       2 2    3 2
ideal(a b*c + a b*d + a c*d*e + a*b*c*d*e + a*b*d e + b d e + c*d e  + d e )
               2
ideal (a + b, b  + d*e)
ideal (e, d, a)
ideal 1
ideal (e, b, a)
ideal (e, c + d, b, a)
                      2
ideal (c + d, a + b, b  + d*e)
ideal (d, a)
ideal (d, c, a)
ideal (d, c, b, a)
ideal (d, b, a)
\end{MyVerbatim}
}
\medskip

>>>>

\medskip
{\small\setstretch{.7}
\begin{MyVerbatim}
i1 : R = ZZ/5[x,y,z];

@@

We can also describe Frobenius roots as follows:  In a (sufficiently local) regular ring, we have an identification of $R$ with its canonical module $\omega_R$.
On the other hand, the Grothendieck dual to the $e$-iterated Frobenius map\footnote{In a reduced ring $R$, it is often convenient to express the Frobenius endomorphism $R \to R$ as the inclusion $R \hookrightarrow R^{1/p}$, so that the source and target  are distinguished.} $R \to R^{1/p^e}$ provides a map
\begin{equation}
\label{eq.DualToFrobenius}
T : \omega_{R^{1/p^e}} \to \omega_R,
\end{equation}
which using the identification $R \cong \omega_R$ gives us a map
\[
T : R^{1/p^e} \to R.
\]
It is not difficult to prove that
$T(I^{1/p^e})=I^{[1/p^e]}$,
where $I^{1/p^e} \subseteq R^{1/p^e}$ is simply the ideal of $p^e$-th roots of the elements of $I$.
In the case that
\[
R = \mathbb{K}[x_1, \ldots, x_d],
\]
where $\mathbb{K}$ is a perfect field,
$R^{1/p^e}$ is a free $R$-module with basis consisting of the monomials
\[
\mathbf{x}^{\boldsymbol{\lambda}/p^e}= x_1^{\lambda_1/p^e} \cdots x_d^{\lambda_d/p^e},
\]
where $0 \leq \lambda_i \leq p^e-1$.
Furthermore, the map $T : R^{1/p^e} \to R$ is simply the projection defined as follows:
\[
T(x_1^{\lambda_1/p^e} \cdots x_d^{\lambda_d/p^e}) = \left\{ \begin{array}{rl} 1 &\text{if } \lambda_i = p^{e}-1 \text{ for all }i \\ 0 & \text{otherwise}  \end{array} \right.
\]
Using this, it is not difficult to see that if
\begin{equation}
\label{eqn1}
f^{1/p^e} = \sum_{\boldsymbol{\lambda}} f_{\boldsymbol{\lambda}} \mathbf{x}^{\boldsymbol{\lambda} / p^e},
\end{equation}
where $\boldsymbol{\lambda}$ runs over the tuples $(\lambda_1, \dots, \lambda_d)$ with $0 \leq \lambda_j \leq p^e-1$, and $f_{\boldsymbol{\lambda}} \in R$, then
$(f)^{[1/p^e]} = T( (f)^{1/p^e})$ is the ideal generated by the $f_{\boldsymbol{\lambda}}$.
We can then compute $I^{[1/p^e]}$ for more general $I$ by linearity: if $I=(f_1,\ldots,f_m)$, then $I^{[1/p^e]}= \sum(f_i)^{[1/p^e]}$.

@@

The computation of Frobenius roots of ideals reduces to the case of principal ideals, and its complexity grows linearly with the number of generators of the ideal.
Furthermore the calculation of  $(f)^{[1/p^e]}$ reduces to finding the terms in the right hand side of (\ref{eqn1}), which essentially amounts to
taking each term in $f$, computing the $p^e$-th root of its coefficient, and dividing the monomial exponent vector by $p^e$ with remainder.
Hence the complexity of computing $(f)^{[1/p^e]}$ is proportional to the number of terms in $f$, and is independent of its degree.
We emphasize that the calculation of Frobenius roots \emph{does not involve the calculation of Gr\"obner bases.}

@@

\medskip
{\small\setstretch{.7}
\begin{MyVerbatim}
i1 : S = ZZ/5[x,y,z];

@@

\medskip
{\small\setstretch{.7}
\begin{MyVerbatim}
i1 : p = 3;

@@

\medskip
{\small\setstretch{.7}
\begin{MyVerbatim}
i8 : S = ZZ/3[a..f,x,y];

@@

\medskip
{\small\setstretch{.7}
\begin{MyVerbatim}
i1 : R = ZZ/2[a,b,c,d];

@@

\medskip
{\small\setstretch{.7}
\begin{MyVerbatim}
i7 : U = matrix entries E2;

@@

i8 : A = image matrix entries relations source E2;

i9 : frobeniusRoot(1, image U)

o9 = image {-2} | 1 0 0 |
           {0}  | 0 d a |

                             2
o9 : R-module, submodule of R
\end{MyVerbatim}
}\medskip

@@

\medskip
{\small\setstretch{.7}
\begin{MyVerbatim}
i10 : B = ascendModule(1, A, U)

o10 = image | 0   a  bc  |
            | c+d d2 a2d |

                              2
o10 : R-module, submodule of R
\end{MyVerbatim}
}\medskip

@@

\medskip
{\small\setstretch{.7}
\begin{MyVerbatim}
i1 : R = ZZ/7[x,y,z]/(x^3 + y^3 + z^3);

@@

\medskip
{\small\setstretch{.7}
\begin{MyVerbatim}
i1 : R = ZZ/7[x,y,z]/((x - 1)^5 + (y + 1)^5 + z^5);

@@

\medskip
{\small\setstretch{.7}
\begin{MyVerbatim}
i1 : R = ZZ/5[x,y,z]/(x^2 + y*z);

@@

\medskip
{\small\setstretch{.7}
\begin{MyVerbatim}
i1 : R = ZZ/5[x,y];

@@

\medskip
{\small\setstretch{.7}
\begin{MyVerbatim}
i1 : R = ZZ/7[x,y,z]/((x - 1)^3 + (y + 1)^3 + z^3);

@@

\medskip
{\small\setstretch{.7}
\begin{MyVerbatim}
i1 : S = ZZ/3[a,b,c,d,t];

@@

\medskip
{\small\setstretch{.7}
\begin{MyVerbatim}
i1 : R = ZZ/5[x,y,z]/(x^4 + y^4 + z^4);

@@

\medskip
{\small\setstretch{.7}
\begin{MyVerbatim}
i1 : R = ZZ/5[x,y,z]/(y*z, x*z, x*y);

@@

\medskip
{\small\setstretch{.7}
\begin{MyVerbatim}
i1 : R = ZZ/2[a..e];

@@

\medskip
{\small\setstretch{.7}
\begin{MyVerbatim}
i1 : T = ZZ/7[x,y];

@@

\medskip
{\small\setstretch{.8}
\begin{MyVerbatim}
i7 : toString QGorensteinGenerator(1, R)

@@

\medskip
{\small\setstretch{.7}
\begin{MyVerbatim}
i1 : T = ZZ/3[x,y];

i2 : S = ZZ/3[a,b,c,d];

@@

Again, consider the maps $T^e : \omega_{R^{1/p^e}} \to \omega_R$ we have considered throughout this section.  It is a theorem of Hartshorne-Speiser, Lyubeznik, and Gabber \cite{HartshorneSpeiserLocalCohomologyInCharacteristicP,LyubeznikFModulesApplicationsToLocalCohomology,Gabber.tStruc} that the descending images
\begin{equation}\label{eq: descending images}
\omega_R \supseteq T(\omega_{R^{1/p}}) \supseteq \dots \supseteq T^e(\omega_{R^{1/p^e}}) \supseteq T^{e+1}(\omega_{R^{1/p^{e+1}}}) \supseteq \cdots
\end{equation}
stabilize for $e \gg 0$.
The function \texttt{FPureModule} computes the stable submodule in this chain, called the \emph{HSLG module}, and returns a sequence containing the following items:
\begin{enumerate}[(1)]
   \item The HSLGmodule;
   \item The canonical module, embedded (non-uniquely) as an ideal;
   \item The element $u$ representing the map on the canonical module (see
   \autoref{subsec.DualToFrobeniusOnQuotientRings});
   \item The value of $e > 0$ at which the descending images in \eqref{eq: descending images} stabilize.  This is sometimes also called the HSLG number of the canonical module as a Cartier module.
\end{enumerate}

\medskip
{\small\setstretch{.7}
\begin{MyVerbatim}
i1 : R = ZZ/3[x,y,z]/(x^3 + y^4 + z^5);

i2 : L = FPureModule R;

i3 : L#0

                          2        2   3
o3 = ideal (0, y*z, x*z, y , x*y, x , z )

i4 : L#1

@@

i5 : L#3

o5 = 1
\end{MyVerbatim}
}\medskip

@@

\medskip
{\small\setstretch{.7}
\begin{MyVerbatim}
i1 : R = ZZ/2[x,y,z];

i2 : f = x^3 + y^2*z + y*z^2;

i3 : frobeniusPower(1/2, ideal f)

o3 = ideal (x, y, z)

i4 : u=f^3;

i5 : frobeniusPower(1/4, ideal u)

o6 = ideal (x, y, z)

@@

o7 : 2
\end{MyVerbatim}
}\medskip

@@

\medskip
{\small\setstretch{.7}
\begin{MyVerbatim}
i1 : R = ZZ/11[x,y,z];

i2 : f = y^2*z^3 - x^5 - 2*z^5;

i3 : frobeniusPower(1/11, ideal f^10)

             2        3
o6 = ideal (z , x*z, x )

@@

             2        3
o6 = ideal (z , x*z, x )

o7 : 2
\end{MyVerbatim}
}\medskip

@@

\medskip
{\small\setstretch{.7}
\begin{MyVerbatim}
i1 : R = ZZ/3[u,v];

i2 : u = u^2*v^2;

i3 : compatibleIdeals u

o3 = {ideal v, ideal (v, u), ideal u}

o3 : List
\end{MyVerbatim}
}\medskip

@@

\medskip
{\small\setstretch{.7}
\begin{MyVerbatim}
i1 : R = ZZ/2[a..f];

i5 : isCohenMacaulay S

@@

\medskip
{\small\setstretch{.7}
\begin{MyVerbatim}
i6 : omega = canonicalIdeal S

o6 = ideal (e, d, a)

o6 : Ideal of S

i7 : omega = substitute(omega, R) + I;

o7 : Ideal of R

i8 : u = intersect((frobenius I):I, (frobenius omega) : omega);

@@

i9 : u = compress((gens u) % gens(frobenius I))

o9 = | a3bc+a3bd+a2cde+abcde+abd2e+b2d2e+cd2e2+d3e2 |

             1       1
o9 : Matrix R  <--- R

i10 : u = first first entries u

       3       3       2                         2     2 2       2 2    3 2
o10 = a b*c + a b*d + a c*d*e + a*b*c*d*e + a*b*d e + b d e + c*d e  + d e

o10 : R
\end{MyVerbatim}
}
\medskip

@@

\medskip
{\small\setstretch{.7}
\begin{MyVerbatim}
i11 : L = compatibleIdeals u;

i12 : print \ L;

        3       3       2                         2     2 2       2 2    3 2
ideal (a b*c + a b*d + a c*d*e + a*b*c*d*e + a*b*d e + b d e + c*d e  + d e )

               2
ideal (a + b, a  + d*e)

ideal (d, a)

ideal (d, a, e)

ideal (e, d, a, c)

ideal (e, d, c, b, a)

ideal (e, d, b, a)

ideal (d, a, c)

ideal (d, c, b, a)

ideal (d, b, a)

                      2
ideal (c + d, a + b, b  + d*e)

ideal (c + d, b, a, e)

ideal (b, a, e)
\end{MyVerbatim}
}
\medskip

@@

\medskip
{\small\setstretch{.7}
\begin{MyVerbatim}
i13 : print \ unique apply(L, J -> J : omega);

        3       3       2                         2     2 2       2 2    3 2
ideal (a b*c + a b*d + a c*d*e + a*b*c*d*e + a*b*d e + b d e + c*d e  + d e )

               2
ideal (a + b, b  + d*e)

ideal (d, a)

ideal (e, d, a)

ideal 1

ideal (d, c, a)

ideal (d, c, b, a)

ideal (d, b, a)

                      2
ideal (c + d, a + b, b  + d*e)

ideal (e, c + d, b, a)

ideal (e, b, a)
\end{MyVerbatim}
}
\medskip
